SFJ - Boolean types and semantic subtyping for Featherweight Java

### 2020-04-06 dissertation/l4proj.tex
final dissertation

\input{semmacros}
\usepackage{amsmath, amssymb}

@@

\title{Boolean types and semantic subtyping for Featherweight Java}
\author{Artem Usov}
\date{January 18, 2020}

@@

\begin{abstract}
    Every abstract follows a similar pattern. Motivate; set aims; describe work; explain results.
    \vskip 0.5em
    ``XYZ is bad. This project investigated ABC to determine if it was better.
    ABC used XXX and YYY to implement ZZZ. This is particularly interesting as XXX and YYY have
    never been used together. It was found that
    ABC was 20\% better than XYZ, though it caused rabies in half of subjects.''
\end{abstract}

@@

\def\consentname {Artem Usov}
\def\consentdate {19 January 2020}
\educationalconsent

@@

\section{End goal of programming}

The end goal in all programming projects is to create a solution to the task at hand that exactly solves the initially defined problem and performs exactly as expected by the programmer.
However, this very rarely actually happens.

@@

In general these are called \emph{bugs} in a program, and have become more and more common as projects and the underlying hardware they run on have become more complex since the early days of computing.
Therefore in both academia and industry there have been great efforts over the years to create tools that allow us to decrease the amount of bugs in our programs and increase our productivity.

@@

One area of great effort has been in the development of static code analysis tools.
Some of these exist integrated into the language itself such as with Spark \citep{Carre1990}.
Spark is a formally defined language based on the Ada \citep{Ada1979} language intended for the development of high integrity software such as flight control systems.
Others exist as separate, well-known tools that exist alongside the language such as the Clang analysis tools for C and C++. \footnote{https://clang.llvm.org/docs/ClangTools.html}

Whilst they are not necessarily needed to write correct programs, these tools have quickly become industry standards for maintaining a level of correctness within codebases and avoiding typical programming errors.

\subsection{How static analysis works}

Underlying how all of the previously mentioned static code analysis tools work, is through their use of the \emph{type system} of the language.
Type systems at their most basic work by assigning types to various constructs of a program so that incompatible types cannot be used.

For example, one common area of error in C code that the Clang tools warn about is implicit conversions, where the value of an expression is a different type from the one expected, such as the use of a floating point number where an integer is expected.
The C language will implicitly convert the value to an integer, however this happens unbeknownst to the programmer, which can cause unexpected behaviour later on in the program.
The tools, using the same knowledge that you can only use an integer where an integer is expected, can therefore flag this as being a potential bug.

@@

The effectiveness of static analysis tools and more generally the \emph{safety} and \emph{expressivity} of a programming language therefore depends on the type system that is used.
Type systems have evolved from the relatively simple systems in languages like C to more complex ones such as in Java which have class types and subtyping relations where types can be substituted for one another.

More modern advances in type systems include the use of linear types in the Cyclone \citep{grossman2002} language that allow us to more naturally and logically define finite state machines.
There also exist other new developments in type systems such as Boolean types which are explored in this paper.

@@

The typical hierarchical subtyping definitions in Java quite often restrict and complicate the logic that the programmer wants to implement.
These problems come up particularly often when working with legacy code that has incorrectly defined abstractions in its classes.
This therefore leads to less understandable and less maintainable code that could cause problems in the future.

@@

This project aims to transform the mathematical definition of such a type system by \citet{Dardha2017} into an actual software implementation, solving the issues encountered doing so.
The language should be able to illustrate improved solutions to the example problems which are encountered in a language without these features such as Java.
Furthermore, the language will be evaluated so see whether these additional language features are intuitive to use.

@@

\begin{itemize}
    \item
          \textbf{Chapter 2} provides background about type systems and Featherweight Java, the language on top of which the new type system is implemented.
          We also discuss the tools which were used in this project and existing work on semantic type systems and boolean types.
    \item
          \textbf{Chapter 3} introduces the design a    nd implementation of SFJ. This includes the syntax of the language, and the key algorithms which were used to implement the type system.
    \item
          \textbf{Chapter 4} TODO
    \item
          \textbf{Chapter 5} TODO
    \item
          \textbf{Chapter 6} contains details about the evaluation method, evaluation results and analysis of results.
    \item
          Finally \textbf{chapter 7} concludes the dissertation with a summary of the important results.
\end{itemize}

@@

In this chapter, we define and explore type systems and the different ways they can be defined. We especially explore the semantic method of defining a type system and explore the benefits and challenges that it brings. We also discuss the tools that were used in this project and the related work that has already been done with boolean types or semantic type systems.

\section{Type Systems}

@@

Most importantly, we want to emphasize from the first definition that in a type system, we aim to classify phrases in our language into a specific \emph{type}.
Then, using the type system, we can check that the types of the arguments given to any operation are always correct, thus avoiding a type error.

For example, taking the addition operation which takes two arguments of the number type, then given the arguments $42$ and $2 + 2$, we know they both hold or compute a number type and so there is guaranteed to be no type error.
Conversely, given the arguments $42$ and the boolean value $true$, they are not both a number type and so cause a type error.
We therefore avoid unexpected bad run-time behaviour and instead find these problems before even running the program.

However what the first definition is missing is that it defines a type system to be solely a syntactic method, in which the system is defined as a list of formal deduction rules.
The syntactic approach is certainly by far the most common approach for a type system, but there does exist an alternative method in the semantic approach.
\citet{Frisch2002} describe the semantic approach as instead starting with a model of the language and all the possible values in the language and defining an interpretation of types as subsets of values the model.

@@

\begin{equation*}
    D = \left[ a + bi \mid a, b \in \mathbb{R}\right]
\end{equation*}

@@

Such as model comes with several advantages.
For example, given two types $s$ and $t$, which represent subsets of values in the model, when $s \subseteq t$ does not hold, it is possible to exhibit an element of the model which is in the interpretation of type $s$ and not $t$.
This can then be used to show more informative error messages to the programmer as shown by \citet{Castagna2005}, such as showing which value in $s$ is causing the equality not to hold.
Secondly the set based approach to the type system leads itself to able to incorporate boolean operators into its types easily.
However due to being a more technical approach, such as the non-triviality of defining the interpretation of types as subsets of a model as discussed below, the semantic method has received less attention than syntactic methods.

The first problem we encounter in the semantic approach is that in such a model, $t_{1}$ is a subtype of $t_{2}$ if all the $t_{1}$-values are also $t_{2}$-values, i.e. the equality $t_{1} \subseteq t_{2}$ holds.
However, in this way, subtyping is defined by relying on the notion of well-typed values; hence, we need the typing relation to able to type values; but the typing rules require a subtyping relation to be fully defined.
So, there is a circularity in our definition, as demonstrated in Figure \ref{fig:circ}.

To solve this problem, we follow the framework defined by \citet{Frisch2008}.
The general idea of the framework is that we first extend the types in the language with \emph{Boolean Combinators}: union $\lor$, intersection $\land$ and negation $\neg$ to give a type algebra $\tau$.
We can then define an abstract model $\beta$ with a interpretation function $[\![ ]\!]_{\beta} : \tau \to \rho(\beta)$ (where $\rho(\beta)$ is the power-set of $\beta$).
This model must capture the meaning of each type but also interprets the combinators in a set-theoretic way, such as $[\![s \land t]\!]_{\beta} = [\![s]\!]_{\beta} \cap [\![t]\!]_{\beta}$ etc.
This model then induces the following subtyping relation:

\begin{equation*}
    s \leq_{\beta} t \iff [\![s]\!]_{\beta} \subseteq [\![t]\!]_{\beta}
\end{equation*}

@@

Separately, now that we have typing rules, we can define a new interpretation of types based on values $[\![t]\!]_{\mathrm{v}} = \left\{\mathrm{v} \in V \mid \: \vdash_{\beta} \mathrm{v} : t \right\}$ and then define a ``new''subtyping relation as above:

\begin{equation*}
    s \leq_{\mathrm{v}} t \iff [\![s]\!]_{\mathrm{v}} \subseteq [\![t]\!]_{\mathrm{v}}
\end{equation*}

While these may be different relations, if the models are chosen carefully such that:
\begin{equation*}
    s \leq_{\beta} t \iff s \leq_{\nu} t,
\end{equation*}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\linewidth]{images/circularity.PNG}
    \caption{Circularity in the semantic subtyping definition \citep{Castagna2005}}
    \label{fig:circ}
\end{figure}

\section{Nominal vs Structural Subtyping}

@@

Nominal subtyping is based on explicit declarations by the developer of the class-hierarchy and is the approach used in Java. $A$ is a subtype of $B$ if and only if it declared to be so by including an explicit declaration that $A$ extends $B$. In Java, every user-defined class either extends another user-defined class or extends the root $Object$ class, creating a tree-like subtyping hierarchy.

Structural subtyping uses the structure of classes, namely its fields and methods, to define the subtyping relation. $A$ is a subtype of $B$ if and only if the fields and methods of $B$ are a subset of the fields and methods of $A$, and that field and method return types are covariant and the method argument types are contravariant. This is similar to the requirements in Liskov's substitution principle \citep{Liskov1994} without the behavioural conditions.

$$
    \begin{array}{ll}
        \syndecl{Apple}{\mathit{Object}} {
         &
            \qquad
            \syndecl{Steel}{\mathit{Object}} {
                \\
                \qquad\quad \mathbf{float}\ \mathit{setPrice}\ (\mathbf{int}\ a)\{\ \ldots\ \}\qquad
         &
                \qquad\qquad\quad \mathbf{int}\ \mathit{quality}\
                \\
            }
            \qquad\quad\qquad
         &
            \qquad\qquad\quad \mathbf{int}\ \mathit{setPrice}\ (\mathbf{float}\ a)\{\ \ldots\ \}
            \\
         &
            \qquad }
    \end{array}
$$

We can see in the example above that $Steel$ is a structural subtype of $Apple$, as we can see that the use of any field or method of $Apple$ would also be valid for an object of type $Steel$. Similarly any field or method return types in $Steel$ are the same or less generic than in $Apple$, and any method argument types are the same or more generic in $Steel$ than in $Apple$. However, $Apple$ is not a subtype of $Steel$ as $Apple$ is missing the $quality$ field. However in some cases a pair of classes can both be subtypes of one another.

@@

In our implementation of SFJ based on the work of \citet{Dardha2017, Dardha2013}, we implement a semantic type system however we include both nominal and structural subtyping as we will further show that this approach gives the programmer a choice in which approach he may wish to use, but also allows greater flexibility and compactness for the logic a programmer may wish to implement.

@@

The syntax of SFJ is exactly that of Featherweight Java (FJ) \citep{Igarashi1999}, with only the associated type system being different.
FJ is closely related to Java, with a key simplification in the removal of assignment operators.
All fields of an object cannot be changed after initialisation and all methods are pure functions.
While this restricts FJ to what is essentially a functional fragment of Java, it is still fully computationally complete.
SFJ was intentionally based on FJ instead of the full Java language because features such as concurrency and reflection are orthogonal to the purpose of demonstrating the implementation of our novel semantic type system.

\paragraph{Multimethods}
While FJ removes some feature of the full Java language, the introduction of boolean types and semantic subtyping restores some of these for free.
Overloaded methods is one of these.
As suggested by \citet{Dardha2017}, we can thus model \emph{multimethods} \citep{BC97}, which according to the authors is ``\emph{very clean and easy to understand [...] it would be the best solution for a brand new language}".
As an example \citet{Dardha2013,Dardha2017} consider the following class declarations:

$$
    \begin{array}{ll}
        \syndecl{A}{\mathit{Object}} {
         &
            \qquad
            \syndecl{B}{A} {
                \\
                \qquad\quad \mathbf{int}\ \mathit{length}\ (\mathbf{string}\ s)\{\ \ldots\ \}\qquad
         &
                \qquad\qquad\quad \mathbf{int}\ \mathit{length}\ (\mathbf{int}\ n)\{\ \ldots\ \}
                \\
            }
         &
            \qquad }
    \end{array}
$$

Method {\it length} has type $\mathbf{string} \rightarrow \mathbf{int}$ in $A$. However, in $B$ it has type $(\mathbf{string} \rightarrow \mathbf{int}) \synwedge (\mathbf{int} \rightarrow \mathbf{int})$,
which can be simplified as $(\mathbf{string} \synvee \mathbf{int}) \rightarrow \mathbf{int}$.

@@

The type system of FSJ was built using ANTLR \citep{parr2013} to define the grammar of the language and automatically create a parser for this grammar rather than having to create one by hand, which made the initial development process much quicker.
It accepts a grammar using Extended Backus-Naur Form (EBNF) \footnote{See ISO/IEC 14977 for reference, although it is not the standard used in the examples} notation to create ANTLR rules which are a list of productions or alternatives.
A general form of a rule is as follow:

$$
    \begin{array}{llll}
        rule & : & alternative_{1}
        \\
             & | & alternative_{2}
        \\
             &   & \vdots
        \\
             & | & alternative_{n}
        \\
             & ; &
    \end{array}
$$

Each alternative production in a rule can itself be a list of elements, where an element can be another rule or just a terminating token.
Since we are using EBNF notation, we can also use \emph{*} and \emph{?} to respectively signify repeated and conditional elements in a production.
We can also use $|$ to give several alternatives for a single element.
We show an excerpt taken from the SFJ grammar as an example:

$$
    \begin{array}{llll}
        expression     & : & \mbox{primExpression ( (PLUS | MINUS | DIV | MULT) primExpression)?}
        \\
                       & ;
        \\
        \\
        primExpression & : & \mbox{NUMBER}
        \\
                       & | & \mbox{TRUE}
        \\
                       & ; &
        \\
        \\
        TRUE           & : & \mbox{'true'}
        \\
                       & ; &
        \\
        \\
        NUMBER         & : & \mbox{DIGIT (DIGIT)*}
        \\
                       & ; &
        \\
        \\
        DIGIT          & : & '0'..'9'
        \\
                       & ; &
    \end{array}
$$

\section{Related Work}
The closest similar area of research would likely be the work on $\mathbb{C}$Duce
\citep{Benzaken2003}, which is xalso a functional language with a semantic type system designed for
working with XML documents and a continuation of the work on XDuce \citep{Hosoya2003}. The language
extended XDuce by introducing less XML specific types such as records, boolean connectives and arrow
types. This therefore makes it similar to (\textit{TODO reword}) our language in that a class-based
semantic type system is a combination of the $\mathbb{C}$Duce record types with arrow types
\textit{(fields and methods)}

Our work and the work on $\mathbb{C}$Duce follow the functional style of $\lambda$-calculus,
whereas the work by \citet{Castagna2008} extends $\pi$-calculus with semantic subtyping.
(\textit{TODO reword}) Similar work to ours creating an implementation for this would result in a
Golang-like \footnote{https://golang.org/} like language, creating a concurrency-focused language with more
intuitive types. The paper found that it was required to be able to decide and resolve the atomicy,
that is whether the only proper subtype is the empty type, of types in order to decide the subtyping
relation, and observes that this same problem appears in $\lambda$-calculus and any other
semantic-based system, for which we will show where we encountered this problem further in this
paper.

@@

% \chapter{Analysis/Requirements}
% What is the problem that you want to solve, and how did you arrive at it?
% \section{Guidance}
% Make it clear how you derived the constrained form of your problem via a clear and logical process.

%===================================================================================================
\chapter{SFJ: Design and Implementation}

\section{Syntax}

@@

\label{sec:design}
The syntax of types is given by the following grammar \citep{Dardha2013,Dardha2017}:
$$
    \begin{array}{llll}
        \tau   & ::=                                       & \alpha \ |\ \mu
               & \mbox{{\em Type term}}
        \\
        \alpha & ::=                                       & \zero \ |\ \btypes \ |\ [\wt{l:\tau}] \ |\ \alpha \ \bf{and}\ \alpha \ |\ \bf{not}\ \alpha
        \qquad
               & \mbox{{\em Object type} ($\alpha $-type)}
        \\
        \mu    & ::=                                       & \alpha \to \alpha \ |\ \mu \ \bf{and}\ \mu \ |\ \bf{not}\ \mu
               & \mbox{{\em Method type} ($\mu $-type)}
    \end{array}
$$

$\alpha$-types are used to type fields and $\mu$-types are used to type methods.
Type $\zero$ is the empty type.
Type $\btypes$ denotes the \emph{basic} types, such as integers, booleans, etc.
    {Record} types $[\wt{l:\tau}]$, where $\wt l$ is a sequence of disjoint labels are used to type objects.
Arrow types $\alpha \to \alpha$ are used to type methods.
The boolean types using ${and}$ and ${not}$ have their expected set-theoretic meanings, and ${or}$ is obtained by their combination.\\

@@

We assume an infinite countable set of names, with some special names: $\mathit{Object}$ indicates the root class, $\synthis$  indicates the current and $\synsuper$ indicates the parent object.
We let  $A, B, C, \ldots$ range over classes; $a, b, \ldots$ over fields; $m, n, \ldots$ over methods and $x, y, z, \ldots$ range over variables.

>>>>

\input{semmacros}
\usepackage{amsmath, amssymb}
\usepackage{float}
\usepackage{changepage}

@@

\title{SFJ - Boolean types and semantic subtyping for Featherweight Java}
\author{Artem Usov}
\date{April 6, 2020}

@@

\begin{abstract}
    The type system of a programming language often dictates its safety and expressivity.
    There are two distinct approaches to defining a type system: the \emph{syntactic} and \emph{semantic}.
    In the semantic approach, one defines a model of the language and an interpretation of types as subsets of this model.
    Subtyping is defined as inclusion of subsets denoting types.
    \vskip 0.5em
    In this paper we present SFJ---Semantic Featherweight Java, an extension of Featherweight Java, which implements such a semantic type system defined by \cite{Dardha2013, Dardha2017}.
    We further exploit it to integrate boolean types as well as nominal and structural subtyping.
    This paper illustrates the benefits of such language constructs by allowing us to write programs that have higher guarantees of correctness by increasing the static type-checking power of the language whilst being more intuitive at the same time.
\end{abstract}

@@

\renewcommand{\abstractname}{Acknowledgements}

\begin{abstract}
    I would like to thank my supervisor Dr Ornela Dardha for her support and guidance throughout the year as well as her enthusiasm for lecturing which is made interested in programming languages and type systems.
    \vskip 0.5em
    I would also like to thank my parents who have always supported and believed in me.
\end{abstract}

\def\consentname {Artem Usov}
\def\consentdate {6 April 2020}
\educationalconsent

@@

\section{End Goal of Programming}

The end goal in all programming projects is to create a solution to the task at hand that exactly solves the initially defined problem and performs exactly as expected by the programmer.
However, this rarely actually happens.

@@

In general these are called \emph{bugs} in a program, and have become more and more common as projects and the underlying hardware they run on have become more complex.
Therefore in both academia and industry there have been great efforts over the years to create tools that allow us to decrease the amount of bugs in our programs and increase our productivity.

@@

One area of great effort has been in the development of static code analysis tools.
Some of these exist integrated into the language itself such as with Spark \citep{Carre1990}.
Spark is a formally defined language based on the Ada \citep{Ada1979} language intended for the development of high integrity software such as flight control systems.
Others exist as separate, well-known tools that exist alongside the language such as the Clang analysis tools for C and C++ \footnote{https://clang.llvm.org/docs/ClangTools.html}.

Whilst they are not necessary to write correct programs, these tools have quickly become industry standards for maintaining a level of correctness within codebases and avoiding typical programming errors.

\subsection{How Static Analysis Works}

Underlying how all of the previously mentioned static code analysis tools work, is their use of the \emph{type system} of the language.
Type systems at their most basic work by assigning types to various constructs of a program so that incompatible types cannot be used.

For example, one common area of error in C code that the Clang tools warn about is implicit conversions, where the value of an expression is a different type from the one expected, such as the use of a floating point number where an integer is expected.
The C language will implicitly convert the value to an integer, however this happens unbeknownst to the programmer, which can cause unexpected behaviour later on in the program.
The tools, using their knowledge of the type system that you can only use an integer where an integer is expected, can therefore flag this as being a potential bug.

@@

The effectiveness of static analysis tools and more generally the \emph{safety} and \emph{expressivity} of a programming language therefore depends on the type system that is uses.
Type systems have evolved from the relatively simple systems in languages like C to more complex ones such as in Java which have class types and subtyping relations where types can be substituted for one another.

More modern advances in type systems include the use of linear types in the Cyclone \citep{grossman2002} language which allows us to more naturally and logically define finite state machines.
There also exist other new developments in type systems such as boolean types which are explored in this paper.

@@

The typical hierarchical subtyping definitions in Java quite often restrict and complicate the logic that the programmer wants to implement.
These problems come up particularly often when working with legacy code which has incorrectly defined abstractions in its classes.
This therefore leads to less understandable and less maintainable code which could cause problems in the future.

@@

This project aims to transform the mathematical definition of such a type system by \citet{Dardha2013, Dardha2017} into an actual software implementation, solving the issues encountered doing so.
The language should be able to illustrate improved solutions to the example problems which are encountered in a language without these features such as Java.
Furthermore, the language will be evaluated so see whether these additional language features are intuitive to use.

@@

\begin{itemize}
    \item
          \textbf{Chapter 2} provides background about type systems and Featherweight Java, the language on top of which the new type system is implemented.
          We also discuss the tools which were used in this project and existing work on semantic type systems and boolean types.
    \item
          \textbf{Chapter 3} introduces the design and implementation of SFJ. This includes the syntax of the language, and the key algorithms which were used to implement the type system.
    \item
          \textbf{Chapter 4} shows how using the semantic type system, we can now easily add language constructs that exploit the set-theoretic implementation of subtyping.
    \item
          \textbf{Chapter 5} explains the process of code generation for a SFJ program so that it can be run on a computer.
    \item
          \textbf{Chapter 6} contains details about the evaluation method, evaluation results and analysis of results.
    \item
          \textbf{Chapter 7} briefly describes the conference paper that was submitted detailing SFJ as a new innovative programming tool.
    \item
          Finally \textbf{chapter 8} concludes the dissertation with a summary of the important results.
\end{itemize}

@@

In this chapter, we explore type systems and the different ways they can be defined.
We especially explore the semantic approach of defining a type system and explore the benefits and challenges that it brings.
We also discuss the tools that were used in this project and the related work that has already been done relating to boolean types or semantic type systems.

\section{Type Systems}
\label{sec:framework}

@@

Most importantly, we want to emphasize from the first definition that in a type system, we aim to classify phrases in our language into a specific \emph{type}.
Then, using the type system, we can check that the types of the arguments given to any operation are always correct, thus avoiding type errors.

For example, taking the addition operation which takes two arguments of the number type, then given the arguments $42$ and $2 + 2$, we know they both hold or compute a number type and so there is guaranteed to be no type error.
Conversely, given the arguments $42$ and the boolean value $true$, they are not both a number type and so would cause a type error.
We therefore avoid unexpected bad run-time behaviour and instead find these problems before even running the program.

However what the first definition is missing is that it defines a type system to be solely a syntactic method, in which the system is defined as a list of formal deduction rules.
The syntactic approach is certainly by far the most common approach for a type system, but there does exist an alternative in the semantic approach.
\citet{Frisch2002} describe the semantic approach as instead starting with a model of the language and all the possible values in the language and defining an interpretation of types as subsets of values the model.

@@

\begin{equation}
    D = \left[ a + bi \mid a, b \in \mathbb{R}\right]
\end{equation}

@@

Such as model comes with several advantages.
For example, given two types $s$ and $t$ which represent subsets of values in the model, when $s \subseteq t$ does not hold then it is possible to exhibit an element of the model which is in the interpretation of type $s$ and not $t$.
This can then be used to show more informative error messages to the programmer as shown by \citet{Castagna2005}, such as showing which value in $s$ is causing the equality not to hold.
Secondly the set based approach to the type system leads itself to able to incorporate boolean operators into its types easily.
However due to being a more technical approach, such as the non-triviality of defining the interpretation of types as subsets of a model discussed below, the semantic method has received less attention than syntactic methods.

The first problem we encounter in the semantic approach is that in such a model $t_{1}$ is a subtype of $t_{2}$ if all $t_{1}$-values are also $t_{2}$-values, i.e. the equality $t_{1} \subseteq t_{2}$ holds.
However, in this way, subtyping is defined by relying on the notion of well-typed values; hence we need the typing relation to able to type values; but the typing rules require a subtyping relation to be fully defined.
So, there is a circularity in our definition, as demonstrated in Figure \ref{fig:circ}.

To solve this problem, we follow the framework defined by \citet{Frisch2008}.
The general idea of the framework is that we first extend the types in the language with \emph{Boolean Combinators}: union $\lor$, intersection $\land$ and negation $\neg$ to give a type algebra $\tau$.
We can then define an abstract model $B$ with a interpretation function $[\![ ]\!]_{B} : \tau \to \rho(B)$ (where $\rho(B)$ is the power-set of $B$).
This model must capture the meaning of each type but also interprets the combinators in a set-theoretic way, such as $[\![s \land t]\!]_{B} = [\![s]\!]_{B} \cap [\![t]\!]_{B}$.
This model then induces the following subtyping relation:

\begin{equation}
    s \leq_{B} t \iff [\![s]\!]_{B} \subseteq [\![t]\!]_{B}
\end{equation}

@@

Separately, now that we have typing rules, we can define a new interpretation of types based on values $[\![t]\!]_{\mathrm{v}} = \left\{\mathrm{v} \in V \mid \: \vdash_{B} \mathrm{v} : t \right\}$ and then define a ``new''subtyping relation as above:

\begin{equation}
    s \leq_{\mathrm{v}} t \iff [\![s]\!]_{\mathrm{v}} \subseteq [\![t]\!]_{\mathrm{v}}
\end{equation}

While these may be different relations, if the models are chosen carefully such that:

\begin{equation}
    s \leq_{B} t \iff s \leq_{\nu} t,
\end{equation}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\linewidth]{images/circularity.PNG}
    \caption{Circularity in trying to define a semantic subtyping relation \citep{Castagna2005}}
    \label{fig:circ}
\end{figure}

\section{Nominal vs Structural Subtyping}
\label{sec:struct}

@@

Nominal subtyping is based on explicit declarations by the developer of the class-hierarchy and is the approach used in Java.
$A$ is a subtype of $B$ if and only if it declared to be so by including an explicit declaration that $A$ extends $B$.
In Java, every user-defined class either extends another class or the root $Object$ class, creating a tree-like subtyping hierarchy.

Structural subtyping uses the structure of classes, namely its fields and methods, to define the subtyping relation.
$A$ is a subtype of $B$ if and only if the fields and methods of $B$ are a subset of the fields and methods of $A$, and if the field and method return types are covariant and the method argument types are contravariant.
This is similar to the requirements in Liskov's Substitution Principle \citep{Liskov1994} without the behavioural conditions.

\begin{equation}
    \label{eq:substitutionprinciple}
    \begin{array}{ll}
        \syndecl{Apple}{\mathit{Object}} {
         &
            \qquad
            \syndecl{Steel}{\mathit{Object}} {
                \\
                \qquad\quad \ldots \qquad
         &
                \qquad\qquad\quad \ldots
                \\
                \qquad\quad \textbf{float}\ \mathit{setPrice}\ (\textbf{int}\ a)\{\ \ldots\ \}\qquad
         &
                \qquad\qquad\quad \textbf{int}\ \mathit{quality}
                \\
            }
            \qquad\quad\qquad
         &
            \qquad\qquad\quad \textbf{int}\ \mathit{setPrice}\ (\textbf{float}\ a)\{\ \ldots\ \}
            \\
         &
            \qquad }
    \end{array}
\end{equation}

We can see in Equation \ref{eq:substitutionprinciple} that $Steel$ is a structural subtype of $Apple$, as the use of any field or method of $Apple$ would also be defined in $Steel$.
All field or method return types in $Steel$ are the same or more restrictive than those that also occur in $Apple$, and all method argument types are the same or less restrictive in $Steel$ than in $Apple$.
However, $Apple$ is not a subtype of $Steel$ as $Apple$ is missing the $quality$ field.
In other cases a pair of classes can both be subtypes of one another.

@@

In our implementation of SFJ based on the work of \citet{Dardha2017, Dardha2013}, we implement a semantic type system however we include both nominal and structural subtyping.
This approach gives the programmer a choice in which approach he may wish to use, but also allows both to be used concurrently for greater flexibility and compactness in the logic a programmer may wish to implement.

@@

The syntax of SFJ is exactly that of Featherweight Java (FJ) \citep{Igarashi1999}, with only the associated type system being different.
FJ is closely related to Java, but with a key simplification in the removal of the assignment operation.
All fields of an object cannot be changed after initialisation and all methods are pure functions.
While this restricts FJ to what is essentially a functional fragment of Java, it is still fully computationally complete.
SFJ was intentionally based on FJ instead of the full Java language because features such as concurrency and reflection are orthogonal to the purpose of demonstrating of our novel type system.

\paragraph{Multimethods}

While FJ removes some feature of the  Java language, the introduction of boolean types and semantic subtyping restores some of these.
Overloaded methods is one of these.
As suggested by \citet{Dardha2017}, we can model overloaded methods as \emph{multimethods} \citep{BC97}, which according to the authors is ``\emph{very clean and easy to understand [...] it would be the best solution for a brand new language}".
As an example \citet{Dardha2013,Dardha2017} consider the following class declarations:

\begin{equation}
    \label{multi1}
    \begin{array}{ll}
        \syndecl{A}{\mathit{Object}} {
         &
            \qquad
            \syndecl{B}{A} {
                \\
                \qquad\quad \ldots \qquad
         &
                \qquad\qquad\quad \ldots
                \\
                \qquad\quad \textbf{int}\ \mathit{length}\ (\textbf{string}\ s)\{\ \ldots\ \}\qquad
         &
                \qquad\qquad\quad \textbf{int}\ \mathit{length}\ (\textbf{int}\ n)\{\ \ldots\ \}
                \\
            }
         &
            \qquad }
    \end{array}
\end{equation}

Method \textit{length} has type $\textbf{string} \rightarrow \textbf{int}$ in $A$.
However, in $B$ it has type $(\textbf{string} \rightarrow \textbf{int}) \synwedge (\textbf{int} \rightarrow \textbf{int})$, which can be simplified as $(\textbf{string} \synvee \textbf{int}) \rightarrow \textbf{int}$.

@@

The type system of FSJ was built using ANTLR \citep{parr2013} to define the grammar of the language and automatically create a parser for this grammar rather than having to create one by hand, which made the initial development process much quicker.
It accepts a grammar using Extended Backus-Naur Form (EBNF) \footnote{See ISO/IEC 14977 for reference, although this is not the standard used in the examples} notation to create ANTLR rules which are a list of productions or alternatives.
A general form of a rule is as follow:

\begin{equation}
    \begin{array}{llll}
        rule & : & alternative_{1}
        \\
             & | & alternative_{2}
        \\
             &   & \vdots
        \\
             & | & alternative_{n}
        \\
             & ; &
    \end{array}
\end{equation}

Each alternative production in a rule can itself be a list of elements, where an element can be another rule or a terminating token.
Since we are using EBNF notation, we can also use $*$ and $?$ to respectively signify repeated and conditional elements in a production.
We can also use $|$ to give several alternatives for a single element.
We show an excerpt taken from the SFJ grammar as an example:

\begin{equation}
    \begin{array}{llll}
        expression     & : & primExpression\ ( (PLUS | MINUS | DIV | MULT)\ primExpression)?
        \\
                       & ;
        \\
        \\
        primExpression & : & NUMBER
        \\
                       & | & TRUE
        \\
                       & ; &
        \\
        \\
        TRUE           & : & 'true'
        \\
                       & ; &
        \\
        \\
        NUMBER         & : & DIGIT\ (DIGIT)*
        \\
                       & ; &
        \\
        \\
        DIGIT          & : & '0'..'9'
        \\
                       & ; &
    \end{array}
\end{equation}

\section{Related Work}
The closest area of research would likely be the work on $\mathbb{C}$Duce
\citep{Benzaken2003}, which is also a functional language with a semantic type system designed for
working with XML documents and a continuation of the work on XDuce \citep{Hosoya2003}.
The language extended XDuce by introducing less XML specific types such as records, boolean connectives and arrow types.
This therefore makes it similar to our language in that a class-based semantic type system is a combination of the $\mathbb{C}$Duce record types with arrow types.
Muehlboeck and Tate \cite{Muehlboeck2018} define a syntactic framework with boolean connectives which has been implemented in the Ceylon programming language \citep{Ceylon2016}.

Our work and the work on $\mathbb{C}$Duce follow the functional style of $\lambda$-calculus, whereas the work by \citet{Castagna2008} extends $\pi$-calculus with semantic subtyping.
Similar work to ours creating an implementation for this would result in a Golang-like \footnote{https://golang.org/} language, creating a concurrency-focused language with more
intuitive types.
\citet{Castagna2008} found that it was required to be able to decide and resolve the atomicy, that is whether the only proper subtype is the empty type, of types in order to decide the subtyping relation, and observes that this same problem appears in $\lambda$-calculus and any other semantic-based system.
This is exactly the problem we find and solve in that we need to validate whether our type definitions are finite trees with basic types as leaves and with not cycles.

@@

\chapter{Syntax and Subtyping Implementation}

This chapter provides the formal syntax of the language and gives a detailed explanation of the implementation of the first core novel feature in SFJ, the semantic type system and subtyping relation.
We then consider the algorithms used in the implementation for their efficiency and any flaws the subtyping relation may have.

\section{Syntax}
\label{sec:syntax}

@@

The syntax of types is given by the following grammar \citep{Dardha2013,Dardha2017}:

\begin{equation}
    \label{eq:syntaxtypes}
    \begin{array}{llll}
        \tau   & ::=                                          & \alpha \ |\ \mu
               & \mbox{\textit{Type term}}
        \\
        \alpha & ::=                                          & \zero \ |\ \btypes \ |\ [\wt{l:\tau}] \ |\ \alpha \ \bf{and}\ \alpha \ |\ \bf{not}\ \alpha
               & \mbox{\textit{Object type} ($\alpha $-type)}
        \\
        \mu    & ::=                                          & \alpha \to \alpha \ |\ \mu \ \bf{and}\ \mu \ |\ \bf{not}\ \mu
               & \mbox{\textit{Method type} ($\mu $-type)}
    \end{array}
\end{equation}

$\alpha$-types are used to type fields and $\mu$-types are used to type methods.
Type $\zero$ is the empty type.
Type $\btypes$ denotes the \emph{basic} types, such as integers, booleans, etc.
Record types $[\wt{l:\tau}]$, where $\wt l$ is a sequence of disjoint labels, are used to type objects.
Arrow types $\alpha \to \alpha$ are used to type methods.
The boolean types using ${and}$ and ${not}$ have their expected set-theoretic meanings, and ${or}$ is obtained by their combination.

@@

We assume an infinite countable set of names, with some special names: $\mathit{Object}$ indicates the root class, $\mathit{this}$  indicates the current and $\mathit{super}$ indicates the parent object.
We let  $A, B, C, \ldots$ range over classes; $a, b, \ldots$ over fields; $m, n, \ldots$ over methods and $x, y, z, \ldots$ range over variables. Constants $c$ range over an infinite countable set $\mathbb{K}$

\begin{equation}
    \label{eq:syntaxterms}
    \begin{array}{llllll}
         & \mbox{\textit{Class declaration}}  & L & \; ::= \; & \syndecl{C}{C}{\wt{\alpha \ a};\ K; \ \wt{M}\ }                         \\
         & \mbox{\textit{Constructor}}        & K & \;::=\;   & C\ (\wt{\alpha\ x})\ \{\ \synsuper(\wt{x});\ \wt{\synthis.a}=\wt{x}; \} \\
         & \mbox{\textit{Method declaration}} & M & \; ::= \; & \alpha | \mu \ m\ (\alpha | \mu \ x)\ \{\ \synret e; \}                 \\
         & \mbox{\textit{Expressions}}        & e & \; ::=\;  & x\ |\  c\ |\ e.a\ |\ e.m(e) \ | \ \synnew C(\wt{e})
    \end{array}
\end{equation}

A program $(\wt{L}, e)$ consists of a sequence of class declarations $\wt L$ and an expression $e$ to be evaluated.
